---
title: "California Fire Project"
output:
  pdf_document: default
  html_document:
    keep_md: yes
  word_document: default
date: '2022-05-10'
---

```{r data-import, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo = TRUE}
##Set your working directory and get your fire data into R
getwd()
setwd("/Users/madikindberg/Desktop/ANS 198")
list.files()

ca_fires2 = read.csv("Fire_Project.csv")

##Make sure your data uploaded correctly and begin to understand columns and rows
ca_fires2
```


library(ggplot2)
library(maps)
library(mapdata)

##Install these packages to begin summarizing the total amount of acres burned 
#in a single county for each individual year

library(tzdb)
library(tidyverse)
library(janitor)
library(glmnet)
library(effects)
library(olsrr)
library(ggrepel)
library(ggtext)
library(glue)
library(dplyr)



##Make these filters in order to subset data for each year to its own table 

ca_fires2_2013 = filter(ca_fires2, Started == 2013)

ca_fires2_2014 = filter(ca_fires2, Started == 2014)

ca_fires2_2015 = filter(ca_fires2, Started == 2015)

ca_fires2_2016 = filter(ca_fires2, Started == 2016)

ca_fires2_2017 = filter(ca_fires2, Started == 2017)

ca_fires2_2018 = filter(ca_fires2, Started == 2018)

ca_fires2_2019 = filter(ca_fires2, Started == 2019)

colnames(ca_fires2_2013)

ca_fires2_2013



```{r echo = FALSE}
##Frequency of fires per year 
#Most fires were shown to be in 2017 

ggplot(ca_fires2, aes(x = factor(Started), fill= AcresBurned)) +
  geom_bar(position = "dodge") +
  labs(title = "Amount of Fires Per Year", x = "Year", y = "Amount of Fires") 

```

```{r echo = TRUE}
##Amount of acres burned per year
ca_all_years_summary = ca_fires2 %>%
  group_by(Started) %>%
  summarise(Sum = sum(AcresBurned, na.rm = TRUE))

ca_all_years_summary 

ggplot(ca_all_years_summary, aes(x = Started, y = Sum)) +
  geom_point() +
  labs(title = "Amount of Acres Burned Per Year", x = "Year", y = "Acres Burned") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = c(2013,2014,2015,2016,2017,2018,2019))
```


##Plot to show all the individual fires in each year
#The plot will also tell us the name of the largest fire in each perspective year
#Purpose of these graphs: What is the largest fire in each year?

```{r echo = TRUE}
#2013: Rim Fire - Tuolumne
ca_fires2_2013

ca_fires2_2013_max = filter(ca_fires2_2013, Name == "Rim Fire")
ca_fires2_2013_max

ca_fires2013_name = ca_fires2_2013_max$Name
ca_fires2013_name

ggplot(ca_fires2_2013) +
  aes(AcresBurned, Counties) +
  geom_point(color = "black", size = 2) +
  geom_point(data = ca_fires2[which.max(ca_fires2_2013$AcresBurned), ], 
  color = "red", size = 2) +
  geom_text_repel(aes(label = ca_fires2013_name), data = ca_fires2_2013_max, size = 5) +
  scale_x_continuous(labels = scales::comma) +
  theme(panel.grid.minor.y = element_line(size = 10))

```




```{r echo = TRUE}
#2014: King Fire - El Dorado
ca_fires2_2014

ca_fires2_2014_max = filter(ca_fires2_2014, Name == "King Fire")
ca_fires2_2014_max

ca_fires2014_name = ca_fires2_2014_max$Name
ca_fires2014_name

ggplot(ca_fires2_2014) +
  aes(AcresBurned, Counties) +
  geom_point(color = "black", size = 2) +
  geom_point(data = ca_fires2_2014[which.max(ca_fires2_2014$AcresBurned), ], 
  color = "red", size = 2) +
  geom_text_repel(aes(label = ca_fires2014_name), data = ca_fires2_2014_max, size = 5) +
  scale_x_continuous(labels = scales::comma) +
  theme(panel.grid.minor.y = element_line(size = 10))
```


```{r echo = TRUE}
#2015: Rough Fire - Fresno 
#why is this not putting out the name on the graph?
ca_fires2_2015

ca_fires2_2015_max = filter(ca_fires2_2015, Name == "Rough Fire")

ca_fires2_2015_max

ca_fires2_2015$Name

ca_fires2015_name = ca_fires2_2015_max$Name
ca_fires2015_name

ggplot(ca_fires2_2015) +
  aes(AcresBurned, Counties) +
  geom_point(color = "black", size = 2) +
  geom_point(data = ca_fires2_2015[which.max(ca_fires2_2015$AcresBurned), ], 
  color = "red", size = 2) +
  geom_text_repel(aes(label = ca_fires2015_name), data = ca_fires2_2015_max, size = 5) +
  scale_x_continuous(labels = scales::comma)  +
  theme(panel.grid.minor.y = element_line(size = 10))
```


```{r echo = TRUE}
#2016 - 
ca_fires2_2016

ca_fires2_2016_max = filter(ca_fires2_2016, Name == "Soberanes Fire")
ca_fires2_2016_max

ca_fires2016_name = ca_fires2_2016_max$Name
ca_fires2016_name

ggplot(ca_fires2_2016) +
  aes(AcresBurned, Counties) +
  geom_point(color = "black", size = 2) +
  geom_point(data = ca_fires2_2016[which.max(ca_fires2_2016$AcresBurned), ], 
  color = "red", size = 2) +
  geom_text_repel(aes(label = ca_fires2016_name), data = ca_fires2_2016_max, size = 5) +
  scale_x_continuous(labels = scales::comma) +
  theme(panel.grid.minor.y = element_line(size = 10))
```


```{r echo = TRUE}
#2017: Thomas Fire - Santa Barbara / Ventura
ca_fires2_2017

ca_fires2_2017_max = filter(ca_fires2_2017, Name == "Thomas Fire")
ca_fires2_2017_max

ca_fires2017_name = ca_fires2_2017_max$Name
ca_fires2017_name

ggplot(ca_fires2_2017) +
  aes(AcresBurned, Counties) +
  geom_point(color = "black", size = 2) +
  geom_point(data = ca_fires2_2017[which.max(ca_fires2_2017$AcresBurned), ], 
  color = "red", size = 2) +
  geom_text_repel(aes(label = ca_fires2017_name), data = ca_fires2_2017_max, size = 5) +
  scale_x_continuous(labels = scales::comma) +
  theme(panel.grid.minor.y = element_line(size = 10))
```


```{r echo = TRUE}
#2018: Ranch Fire - Large enough for multiple Counties
ca_fires2_2018_max = filter(ca_fires2_2018, Name == "Ranch Fire (Mendocino Complex)")
ca_fires2_2018_max

ca_fires2018_name = ca_fires2_2018_max$Name
ca_fires2018_name

ggplot(ca_fires2_2018) +
  aes(AcresBurned, Counties) +
  geom_point(color = "black", size = 2) +
  geom_point(data = ca_fires2_2018[which.max(ca_fires2_2018$AcresBurned), ], 
  color = "red", size = 2) +
  geom_text_repel(aes(label = ca_fires2018_name), data = ca_fires2_2018_max, size = 5) +
  scale_x_continuous(labels = scales::comma) +
  theme(panel.grid.minor.y = element_line(size = 10))
```


```{r echo = TRUE}
#2019
ca_fires2_2019_max = filter(ca_fires2_2019, Name == "Kincade Fire")
ca_fires2_2019_max

ca_fires2019_name = ca_fires2_2019_max$Name
ca_fires2019_name

ggplot(ca_fires2_2019) +
  aes(AcresBurned, Counties) +
  geom_point(color = "black", size = 2) +
  geom_point(data = ca_fires2_2019[which.max(ca_fires2_2019$AcresBurned), ], 
  color = "red", size = 2) +
  geom_text_repel(aes(label = ca_fires2019_name), data = ca_fires2_2019_max, size = 5) +
  scale_x_continuous(labels = scales::comma) +
  theme(panel.grid.minor.y = element_line(size = 10))
```



##Start summarizing the total amount of acres burned in a single county for all counties in each individual year
```{r echo = TRUE}
#Summary for 2013
ca_2013 = ca_fires2_2013 %>%
  group_by(Counties) %>%
  summarise(Sum = sum(AcresBurned))
  
ca_2013

#Summary for 2014
ca_2014 = ca_fires2_2014 %>%
  group_by(Counties) %>%
  summarise(Sum = sum(AcresBurned))
  
ca_2014

#Summary for 2015
ca_2015 = ca_fires2_2015 %>%
  group_by(Counties) %>%
  summarise(Sum = sum(AcresBurned))
  
ca_2015

#Summary for 2016
ca_2016 = ca_fires2_2016 %>%
  group_by(Counties) %>%
  summarise(Sum = sum(AcresBurned))
  
ca_2016

#Summary for 2017
ca_2017 = ca_fires2_2017 %>%
  group_by(Counties) %>%
  summarise(Sum = sum(AcresBurned))
  
ca_2017

#Summary for 2018
ca_2018 = ca_fires2_2018 %>%
  group_by(Counties) %>%
  summarise(Sum = sum(AcresBurned))
  
ca_2018

#Summary for 2019
ca_2019 = ca_fires2_2019 %>%
  group_by(Counties) %>%
  summarise(Sum = sum(AcresBurned))
  
ca_2019
```


##Start outlining maps and learning how to use the map_data
```{r echo = TRUE}
#For starters, this will show us the map of California

usa = map_data("usa")
states= map_data("state")
dim(states)

ca_df= subset(states, region == "california")

ca_base= ggplot(data=ca_df, mapping= aes(x=long, y=lat, group=group)) +
  coord_fixed(1.3) +
  geom_polygon(color = "black", fill = "gray")

```


```{r echo = TRUE}
#Now explore showing the counties of California on the map
counties = map_data("county")

ca_county = subset(counties, region== "california")

ca_base +
  geom_polygon(data = ca_county, fill = NA, color = "white") +
  geom_polygon(color = "black", fill=NA)
```


```{r echo = TRUE}
##Let's focus on plotting our data on the California counties map now
library(stringr)
library(dplyr)

#This will give us a shortcut to Acres Burned data we will focus on 
fire_per_mile = ca_fires2$AcresBurned

fire_per_mile

##Tolower will allow us to lowercase our "Counties" names in our data in order 
#to align it will the California map data that is also lowercase which allows 
#us to join and plot our data on the map
#Use this for each year so that all our plots will join easily with the map
ca_fires2$Counties2 = tolower(ca_fires2$Counties)
ca_fires2$Counties2

ca_2013$Counties2013 = tolower(ca_2013$Counties)
ca_2013$Counties2013

ca_2014$Counties2014 = tolower(ca_2014$Counties)
ca_2014$Counties2014

ca_2015$Counties2015 = tolower(ca_2015$Counties)
ca_2015$Countixes2015

ca_2016$Counties2016 = tolower(ca_2016$Counties)
ca_2016$Counties2016

ca_2017$Counties2017 = tolower(ca_2017$Counties)
ca_2017$Counties2017

ca_2018$Counties2018 = tolower(ca_2018$Counties)
ca_2018$Counties2018

ca_2019$Counties2019 = tolower(ca_2019$Counties)
ca_2019$Counties2019

##Put description here!! Why do we need elbow room??
#Put description here!! Why do we need ditch the axis??

ditch_the_axes = theme(axis.text = element_blank(),
                       axis.line = element_blank(),
                       axis.ticks = element_blank(),
                       panel.border = element_blank(),
                       panel.grid = element_blank(),
                       axis.title = element_blank()
)

elbow_room1 = ca_base +
  geom_polygon(data = ca_fires2, aes(fill = fire$AcresBurned), color = "white") +
  geom_polygon(color = "black", fill= NA) +
  theme_bw() +
  ditch_the_axes

```


```{r echo = TRUE}
##Build the base for the maps
ca_countymap = ca_base +
  geom_polygon(data = ca_county, fill = NA, color = "white") +
  geom_polygon(color = "black", fill=NA)
ca_countymap
```


```{r echo = TRUE}
#Now we can finally join our data with the California map data 
#Essentially this will overlap our columns for counties in each set of data
#For the California map data, the counties are found under "subregion"
#For our fire data, the counties are found under "CountiesYear"
ca_countymap2 = ca_county %>%
  left_join(ca_fires2, by = c("subregion" = "Counties2"))
ca_countymap2

ca_countymap2013 = ca_county %>%
  left_join(ca_2013, by = c("subregion" = "Counties2013"))
ca_countymap2013

ca_countymap2014 = ca_county %>%
  left_join(ca_2014, by = c("subregion" = "Counties2014"))
ca_countymap2014

ca_countymap2015 = ca_county %>%
  left_join(ca_2015, by = c("subregion" = "Counties2015"))
ca_countymap2015

ca_countymap2016 = ca_county %>%
  left_join(ca_2016, by = c("subregion" = "Counties2016"))
ca_countymap2016

ca_countymap2017 = ca_county %>%
  left_join(ca_2017, by = c("subregion" = "Counties2017"))
ca_countymap2017

ca_countymap2018 = ca_county %>%
  left_join(ca_2018, by = c("subregion" = "Counties2018"))
ca_countymap2018

ca_countymap2019 = ca_county %>%
  left_join(ca_2019, by = c("subregion" = "Counties2019"))
ca_countymap2019
```


```{r echo = TRUE}
#Making the intial graph for all years
acresmap = ca_base +
  geom_polygon(data = ca_countymap2,  aes(fill = AcresBurned),
               color = "white") +
  geom_polygon(color = "black", fill = NA) +
  theme_bw() +
  ditch_the_axes
acresmap

#Adding layers to our intial map for all years
acresmap2 = acresmap + 
  ggtitle("Acres Burned in California 2013-2019") +
  scale_fill_gradientn(colors = rev(rainbow(7)),
                      breaks = c(10, 100, 1000, 10000, 250000),
                      trans = "log10")
acresmap2
```


```{r echo = TRUE}
#Making the initial graph for 2013
acresmap2013 = ca_base +
  geom_polygon(data = ca_countymap2013,  aes(fill = Sum),
               color = "white") +
  geom_polygon(color = "black", fill = NA) +
  theme_bw() +
  ditch_the_axes
acresmap2013

#Adding layers to our intial map for 2013
acresmap2013B = acresmap2013 + 
  ggtitle("Acres Burned in California 2013") +
  scale_fill_gradientn(colors = rev(rainbow(7)),
                      breaks = c(10, 100, 1000, 10000, 250000),
                      trans = "log10")
acresmap2013B
```


```{r echo = TRUE}
#Making the initial graph for 2014
acresmap2014 = ca_base +
  geom_polygon(data = ca_countymap2014,  aes(fill = Sum),
               color = "white") +
  geom_polygon(color = "black", fill = NA) +
  theme_bw() +
  ditch_the_axes
acresmap2014

#Adding layers to our intial map for 2014
acresmap2014B = acresmap2014 + 
  ggtitle("Acres Burned in California 2014") +
  scale_fill_gradientn(colors = rev(rainbow(7)),
                      breaks = c(10, 100, 1000, 10000, 250000),
                      trans = "log10")
acresmap2014B
```


```{r echo = TRUE}
#Making the initial graph for 2015
acresmap2015 = ca_base +
  geom_polygon(data = ca_countymap2015,  aes(fill = Sum),
               color = "white") +
  geom_polygon(color = "black", fill = NA) +
  theme_bw() +
  ditch_the_axes
acresmap2015

#Adding layers to our intial map for 2015
acresmap2015B = acresmap2015 + 
  ggtitle("Acres Burned in California 2015") +
  scale_fill_gradientn(colors = rev(rainbow(7)),
                      breaks = c(10, 100, 1000, 10000, 250000),
                      trans = "log10")
                      
acresmap2015B
```


```{r echo = TRUE}
#Making the initial graph for 2016
acresmap2016 = ca_base +
  geom_polygon(data = ca_countymap2016,  aes(fill = Sum),
               color = "white") +
  geom_polygon(color = "black", fill = NA) +
  theme_bw() +
  ditch_the_axes
acresmap2016

#Adding layers to our intial map for 2016
acresmap2016B = acresmap2016 + 
  ggtitle("Acres Burned in California 2016") +
  scale_fill_gradientn(colors = rev(rainbow(7)),
                      breaks = c(10, 100, 1000, 10000, 250000),
                      trans = "log10")
acresmap2016B
```


```{r echo = TRUE}
#Making the initial graph for 2017
acresmap2017 = ca_base +
  geom_polygon(data = ca_countymap2017,  aes(fill = Sum),
               color = "white") +
  geom_polygon(color = "black", fill = NA) +
  theme_bw() +
  ditch_the_axes
acresmap2017

#Adding layers to our intial map for 2017
acresmap2017B = acresmap2017 + 
  ggtitle("Acres Burned in California 2017") +
  scale_fill_gradientn(colors = rev(rainbow(7)),
                      breaks = c(10, 100, 1000, 10000, 250000),
                      trans = "log10")
acresmap2017B
```


```{r echo = TRUE}
#Making the initial graph for 2018
acresmap2018 = ca_base +
  geom_polygon(data = ca_countymap2018,  aes(fill = Sum),
               color = "white") +
  geom_polygon(color = "black", fill = NA) +
  theme_bw() +
  ditch_the_axes
acresmap2018

#Adding layers to our intial map for 2018
acresmap2018B = acresmap2018 + 
  ggtitle("Acres Burned in California 2018") +
  scale_fill_gradientn(colors = rev(rainbow(7)),
                      breaks = c(10, 100, 1000, 10000, 250000),
                      trans = "log10")
acresmap2018B
```


```{r echo = TRUE}
#Making the initial graph for 2019
acresmap2019 = ca_base +
  geom_polygon(data = ca_countymap2019,  aes(fill = Sum),
               color = "white") +
  geom_polygon(color = "black", fill = NA) +
  theme_bw() +
  ditch_the_axes
acresmap2019

#Adding layers to our intial map for 2019
acresmap2019B = acresmap2019 + 
  ggtitle("Acres Burned in California 2019") +
  scale_fill_gradientn(colors = rev(rainbow(7)),
                      breaks = c(10, 100, 1000, 10000, 250000),
                      trans = "log10")
acresmap2019B
```

```{r echo = TRUE}
##Now move onto analyzing relationship between California fires and rainfall
##Start by loading all packages needed
library(ggplot2)

library(maps)
library(mapdata)
library(tidyverse)
library(janitor)
library(glmnet)
library(effects)
library(olsrr)
library(ggrepel)
library(ggtext)
library(glue)
library(dplyr)
```


```{r echo = TRUE}
##Load all the rainfall datasets we will need to use
getwd()
setwd("/Users/madikindberg/Desktop/ANS 198")
list.files()

rainfall_2013 = read.csv("Rainfall 2013.csv")

rainfall_2014 = read.csv("Rainfall 2014.csv")

rainfall_2015 = read.csv("Rainfall 2015.csv")

rainfall_2016 = read.csv("Rainfall 2016.csv")

rainfall_2017 = read.csv("Rainfall 2017.csv")

rainfall_2018 = read.csv("Rainfall 2018.csv")

rainfall_2019 = read.csv("Rainfall 2019.csv")




rainfall_2013_summary = rainfall_2013 %>%
  group_by(County) %>%
  summarise(Sum = sum(Year.Total))

rainfall_2013_summary

sum(rainfall_2013$Year.Total)
sum(rainfall_2014$Year.Total)
sum(rainfall_2015$Year.Total)
sum(rainfall_2016$Year.Total)
sum(rainfall_2017$Year.Total)
sum(rainfall_2018$Year.Total)
sum(rainfall_2019$Year.Total)
```


```{r echo = TRUE}
##Load the fire data
getwd()
setwd("/Users/madikindberg/Desktop/ANS 198")
list.files()

ca_fires2 = read.csv("Fire_Project.csv")

ca_fires2_2013 = filter(ca_fires2, Started == 2013)
ca_fires2_2014 = filter(ca_fires2, Started == 2014)
ca_fires2_2015 = filter(ca_fires2, Started == 2015)
ca_fires2_2016 = filter(ca_fires2, Started == 2016)
ca_fires2_2017 = filter(ca_fires2, Started == 2017)
ca_fires2_2018 = filter(ca_fires2, Started == 2018)
ca_fires2_2019 = filter(ca_fires2, Started == 2019)

sum(ca_fires2_2013$AcresBurned)
sum(ca_fires2_2014$AcresBurned)
sum(ca_fires2_2015$AcresBurned)
sum(ca_fires2_2016$AcresBurned)
sum(ca_fires2_2017$AcresBurned)
sum(ca_fires2_2018$AcresBurned)
sum(ca_fires2_2019$AcresBurned, na.rm = TRUE)


nrow(ca_fires2_2013)
nrow(ca_fires2_2014)
nrow(ca_fires2_2015)
nrow(ca_fires2_2016)
nrow(ca_fires2_2017)
nrow(ca_fires2_2018)
nrow(ca_fires2_2019)
```

#Make a dataset with summaries for fire and rainfall and use in next portion

```{r echo = TRUE}
##Read the data with rainfall and acres burned
ca_total_data = read.csv("Rainfall from 2013-2019.csv")
ca_total_data

Acres_Burned = ca_total_data$Acres.Burned
Acres_Burned 

Total_Rainfall = ca_total_data$Total.Rainfall..inches.
Total_Rainfall

All_Years = ca_total_data$Year
All_Years
```


```{r echo = TRUE}
##Make a graph with two separate axis
par(mar = c(5,4,4,4)+0.3)

plot(All_Years, Acres_Burned, pch = 16, col = 2, xlab = "Year", 
  ylab = "Acres Burned") 

par(new = TRUE)

plot(All_Years, Total_Rainfall, pch = 17, col = 3,
     axes = FALSE, xlab = "", ylab = "") 

axis(side = 4, at = pretty(range(Total_Rainfall)))

mtext("Total Rainfall (inches)", side = 4, line = 3)

mtext("Trend in Acres Burned and Rainfall from 2013 to 2019", side = 3, 
  line = 1)

legend("topleft", inset = c(0, 0), legend = c("Acres Burned", "Total Rainfall"),
       text.col = c("red","green"), pch = c(1,4), col = c("red","green"), 
       title = "Key")
```


